# Reproducible Research course project week 2
## Adam Roscoe

## Loading and preprocessing the data

```{r, echo=FALSE, results='hide', warningFALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```

### Download and unzip

```{r, results='markup', warning=TRUE, message=TRUE}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data/activity.zip",method="curl")
```

```{r}
unzip(zipfile="./data/activity.zip",exdir="./data")
activity <- read.csv("./data/activity.csv")
activity$date <- as.Date(activity$date)
```

### Calculate the total number of steps taken per day.

```{r}
stepsPerDay <- activity %>%
    group_by(date) %>%
    summarize(sumsteps = sum(steps, na.rm = TRUE))
head(stepsPerDay,10)
```

```{r}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="green", xlab="Steps", ylim = c(0,30))
```

### Calculate mean

```{r}
meanSteps <- round(mean(stepsPerDay$sumsteps),digits = 2)
medianSteps <- round(median(stepsPerDay$sumsteps),digits = 2)
print(paste("The mean is: ", meanSteps))
```
### Calculate median

```{r}
print(paste("The median is: ", medianSteps))
```

### Average of daily activity pattern

```{r}
stepsPerInterval <- activity %>%
    group_by(interval) %>%
    summarize(meansteps = mean(steps, na.rm = TRUE)) 
#Display first 10 rows of data
```

```{r}
head(stepsPerInterval,10)
```

```{r}
plot(stepsPerInterval$meansteps ~ stepsPerInterval$interval,
     col="green", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")
```

### Which 5- minute interval contains the maximum number of steps?
```{r}
print(paste("Interval containing the most steps on average: ",stepsPerInterval$interval[which.max(stepsPerInterval$meansteps)]))
```

```{r}
print(paste("Average steps for that interval: ",round(max(stepsPerInterval$meansteps),digits=2)))
```

### Imputing missing values
``{r}
print(paste("The total number of rows with NA is: ", sum(is.na(activity$steps))))
```

### NA strategy: to solve for the missing NA values the average for the associated interval will be used.
### 1. loop through records of a copy of 'activity' data.
### 2. look for records containing NA values
### 3. transform the 'steps' value based on matching the interval.
```{r}
head(activity, 10)
```
```{r}
activityNoNA <- activity
for (i in 1: nrow(activity)){
        if(is.na(activity$steps[i])){
                activityNoNA$steps[i] <- stepsPerInterval$meansteps[activityNoNA$interval[i] == stepsPerInterval$interval]
        }
}
```

```{r}
head(activityNoNA, 10)
```
### Histogram for total number of steps taken each day

```{r}
stepsPerDay <- activityNoNA %>%
    group_by(date) %>%
    summarize(sumsteps = sum(steps, na.rm = TRUE))
head(stepsPerDay, 10)
```

```{r}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps",
     col = "red", xlab= "Steps")
```
### mean post NA

```{r}
meanPostNA <- round(mean(stepsPerDay$sumsteps), digits = 2)
medianPostNA <- round(median(stepsPerDay$sumsteps), digits = 2)
```

```{r}
print(paste("the mean is: ", mean(meanPostNA)))
```

```{r}
print(paste("The median is : ", median(medianPostNA)))
```

### Difference in activity patterns between weekdays and weekends

```{r}
activityDoW <- activityNoNA
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)
```

```{r}
activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

activityWeekday <- activityWeekday %>%
    group_by(interval) %>%
    summarize(steps = mean(steps)) 
activityWeekday$day <- "weekday"

activityWeekend <- activityWeekend %>%
    group_by(interval) %>%
    summarize(steps = mean(steps)) 
activityWeekend$day <- "weekend"

wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)


g <- ggplot (wkdayWkend, aes (interval, steps))
g + geom_line() + facet_grid (day~.) + 
    theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
    labs(y = "Number of Steps") + labs(x = "Interval") + 
    ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
    theme(plot.title = element_text(hjust = 0.5))
```